High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-04-25 notebooks/mnist.qmd
work on mnist example

```{julia}
# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.9
T = 50
```

```{julia}
# Search:
opt = Flux.Optimise.Adam(0.01)
generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
```

@@

```{julia}
# Define generator:
generator = REVISEGenerator(
  opt = opt,
  λ=0.01
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x, target, counterfactual_data, M, generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
```

@@

```{julia}
# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive)
    conf_model = conformal_model(jem_ens; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    fit!(mach)
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
model_dict = Dict(mod_name => _train(mod) for (mod_name, mod) in models)
```

```{julia}
if mach.model.model isa JointEnergyClassifier
    sampler = mach.model.model.jem.sampler
else
    K = length(counterfactual_data.y_levels)
    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
    𝒟x = Uniform(extrema(counterfactual_data.X)...)
    𝒟y = Categorical(ones(K) ./ K)
    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
end
opt = ImproperSGLD()
f(x) = logits(M, x)

n_iter = 200
_w = 1500
plts = []
neach = 10
for i in 1:10
    x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
    plts_i = []
    for j in 1:size(x, 2)
        xj = x[:,j]
        xj = reshape(xj, (n_digits, n_digits))
        plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
    end
    plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
    plts = [plts..., plt]
end
plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
display(plt)

```

```{julia}
test_data = load_mnist_test()
test_data.X = pre_process.(test_data.X)
f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
println("F1 score (test): $(round(f1,digits=3))")
```

@@

# Set up search:
factual_label = 9
x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 7
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.9
T = 100

@@

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(0.01),)
ce_wachter = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

generator = GreedyGenerator(η=η)
ce_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

@@

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(0.01),
)
ce_conformal = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
)
ce_conformal_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "eccco_mnist.png"))
```

@@

# Set up search:
factual_label = 8
x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.5
T = 100

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
ce_wachter = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

@@

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(),
)
ce_conformal = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
)
ce_conformal_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

>>>>

# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.9

# Training params:
T = 100
opt = Flux.Optimise.Adam(0.01)
```

```{julia}
# Search:
generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "you_may_not_like_it.png"))
```

@@

```{julia}
# Define generator:
generator = REVISEGenerator(
  opt = opt,
  λ=0.1
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
```

@@

```{julia}
# Hyper:
_retrain = false
_regen = false

# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
    conf_model = conformal_model(model; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    @info "Begin training $mod_name."
    fit!(mach)
    @info "Finished training $mod_name."
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
if _retrain
    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
    Serialization.serialize(joinpath(output_path,"mnist_models.jls"), model_dict)
else
    model_dict = Serialization.deserialize(joinpath(output_path,"mnist_models.jls"))
end
```

```{julia}
# Plot generated samples:
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

        n_iter = 200
        _w = 1500
        plts = []
        neach = 10
        for i in 1:10
            x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
            plts_i = []
            for j in 1:size(x, 2)
                xj = x[:,j]
                xj = reshape(xj, (n_digits, n_digits))
                plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
            end
            plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
            plts = [plts..., plt]
        end
        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
        savefig(plt, joinpath(output_images_path, "mnist_generated_$(mod_name).png"))
        display(plt)
    end
end
```

```{julia}
# Evaluate models:

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    test_data = load_mnist_test()
    test_data.X = pre_process.(test_data.X, noise=0.0f0)
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
model_performance
```

```{julia}
# ECCCo:
λ=[0.5,0.1,0.5]
temp=0.1
η=0.01

# Generate counterfactuals using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(η),
)

ces = Dict()
for (mod_name, mod) in model_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, mod, generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )
    ces[mod_name] = ce
end
plt_order = sortperm(collect(keys(ces)))

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)

plts = []
for (_name,ce) in ces
    _x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plts = plts[plt_order]
plts = [p1, plts...]
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
```

@@

# Set up search:
factual_label = 9
x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 7
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.9
T = 100

@@

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(0.01),)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

generator = GreedyGenerator(η=η)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

@@

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(0.01),
)
ce_conformal = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
)
ce_conformal_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x_factual = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x_factual,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "eccco_mnist.png"))
```

@@

# Set up search:
factual_label = 8
x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.5
T = 100

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

@@

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(),
)
ce_conformal = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
)
ce_conformal_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]
